Bird beak's elaboration and exploration: data preparation

### 2022-08-23 Analysis/01-Data_preparation.Rmd
TODO!
Update 01-Data_preparation.Rmd

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
The following code snippets are used to illustrate the implementation of the functions.
See §§§ for running the actual analyses.

The method described in the next three markdown is essentially a mini chains with pseudo-metropolis coupling markov chain generalised linear mixed model method to determine clades and tips elaboration and exploration (`mcmcpmcmcmcglmmmctee`) but we will refrain using this acronym and method name.

```{r install_beer, eval = FALSE}
## Installing the dispRity branch
devtools::install_github("TGuillerme/dispRity@MCMCglmm")
```

```{r load_beer, eval = TRUE}
## Loading packages
library(dispRity) #v>=1.7
library(xtable)
library(knitr)
source("../Functions/stats.utilities.R")

```

@@

TODO:

 * [ ] what's the data, where is it from
 * [ ] Procrustes
 * [ ] Building the trait space
   - [ ] form space version?
   - [ ] shape space version?
 * [ ] Clade data and grouping

@@

 * The clade data
We used the different clade levels (the 11 magnificent 7 + the families). For each level/group, we removed any groups with less than 20 species.

@@

```{r}
## Get the traitspace for all passeriformes
load("../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
shapespace_passeriformes <- shapespace_allbirds_lvl_superorder_order[[1]]$space[(clades_list[, "level2_order"] == "Passeriformes"), ]

## Check what's happening with Neodrepanis_hypoxantha

@@

## Fill in the level1
for(clade in 1:length(clade_info_passerines$passerine_big_clades)) {
  ## Get the clade name
  clade_name <- names(clade_info_passerines$passerine_big_clades)[clade]
  ## Attribute the species/clade
  shapespace_passeriformes[clade_info_passerines$passerine_big_clades[[clade]], "level1"] <- clade_name
  ## Remove species not present originally in the shapespace
  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
  }
}
## Fill in the level2
for(clade in 1:length(clade_info_passerines$passerine_small_clades)) {
  ## Get the clade name
  clade_name <- names(clade_info_passerines$passerine_small_clades)[clade]
  ## Attribute the species/clade
  shapespace_passeriformes[clade_info_passerines$passerine_small_clades[[clade]], "level2"] <- clade_name
  ## Remove species not present originally in the shapespace
  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
  }
}
shapespace_passeriformes[, "level1"] <- as.factor(shapespace_passeriformes[, "level1"])
shapespace_passeriformes[, "level2"] <- as.factor(shapespace_passeriformes[, "level2"])

## Do the prep data from here (with all the trees! ouch!)
shapespace_passeriformes_two_levels <- prep.data(
                                          level = 0, lvl.inc = c(1,2),
                                          clades    = shapespace_passeriformes[, c("level1", "level2")],
                                          space     = shapespace_passeriformes,
                                          dim       = 1:8,
                                          consensus = consensus_tree,
                                          trees     = trees_list,
                                          verbose   = TRUE)
save(shapespace_passeriformes_two_levels, file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
```

>>>>

All the custom code and custom functions called in this document can be downloaded and installed in `R`.

```{r packages, eval = TRUE}
## Loading packages
library(dispRity) #v>=1.7
library(xtable)
library(knitr)
source("../Functions/stats.utilities.R")
```

@@

@@

 * The clade data
We used the different clade levels (super orders, orders, sub-orders and families). For each level/group, we removed any groups with less than 20 species.

@@

```{r}
source("../Functions/prep.data.R")
## Get the traitspace for all passeriformes
load("../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
shapespace_passeriformes <- shapespace_allbirds_lvl_superorder_order[[1]]$space[shapespace_allbirds_lvl_superorder_order[[1]]$space$level2 == "Passeriformes", ]

@@

## Fill in the level1
for(clade in 1:length(clade_info_passerines$passerine_big_clades)) {
  ## Get the clade name
  clade_name <- names(clade_info_passerines$passerine_big_clades)[clade]
  ## Attribute the species/clade
  shapespace_passeriformes[clade_info_passerines$passerine_big_clades[[clade]], "level1"] <- clade_name
  ## Remove species not present originally in the shapespace
  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
  }
}

## Fill in the level2
for(clade in 1:length(clade_info_passerines$passerine_small_clades)) {
  ## Get the clade name
  clade_name <- names(clade_info_passerines$passerine_small_clades)[clade]
  ## Attribute the species/clade
  shapespace_passeriformes[clade_info_passerines$passerine_small_clades[[clade]], "level2"] <- clade_name
  ## Remove species not present originally in the shapespace
  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
  }
}
shapespace_passeriformes[, "level1"] <- as.factor(shapespace_passeriformes[, "level1"])
shapespace_passeriformes[, "level2"] <- as.factor(shapespace_passeriformes[, "level2"])

## Do the prep data from here
shapespace_passeriformes_two_levels <- prep.data(
                                          level = 0, lvl.inc = c(1,2),
                                          clades    = shapespace_passeriformes[, c("level1", "level2")],
                                          space     = shapespace_passeriformes,
                                          dim       = 1:8,
                                          consensus = consensus_tree,
                                          trees     = trees_list,
                                          verbose   = TRUE)
save(shapespace_passeriformes_two_levels, file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
```
